---
title: "GSE145491 CITE-Seq Analysis"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

# Set Up

### Loading in required packages
```{r libraries, results='hide'}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(biomaRt)
library(pheatmap)
library(fgsea)
library(patchwork)
options(future.globals.maxSize = 4000 * 1024^2)
setwd("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491")

run_gsea <- function(markers, gmt_path, output_gsea) {
rnk <- setNames(markers$avg_log2FC, rownames(markers))
rnk <- rnk[!is.na(rnk)]
gmt <- gmtPathways(gmt_path)
gsea <- fgsea(pathways = gmt, stats = rnk, minSize = 15, maxSize = 500)
gseadf <- as.data.frame(gsea)
list_columns <- sapply(gseadf, is.list)
gseadf[list_columns] <- lapply(gseadf[list_columns], function(col) sapply(col, paste, collapse = ";"))
write.csv(gseadf, output_gsea)
}

EnsIDReplace2 <- function(input, output){
  gene_mapping <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/Ensembl_labels.csv")
  rna <- input
  rnanames <- rownames(rna)
  id_list <- setNames(gene_mapping$gene_name, gene_mapping$gene_id)
  rnaname <- character(nrow(rna))
  for (i in seq_len(nrow(rna))) {
    ensembl_id <- rownames(rna)[i]
    if (ensembl_id %in% names(id_list)) {
      rnaname[i] <- id_list[[ensembl_id]]
    } else{
      rnaname[i] <- ensembl_id
    }
  }
  rnaname[rnaname == ""] <- NA
  
  rownames(rna) <- rnaname
  output <- rna
  rm(gene_mapping, id_list, ensembl_id, i, rnaname)
}
```

Load the dataset and create Seurat Object with two assays

```{r load-data}
data <- as.sparse(read.table(file = "GSE145491_UMIsMatrix.txt.gz",
                               sep = "", header = TRUE, row.names = 1))
adtfeat <- c("CD9", "CD41", "CD48", "CD55", "CD105", "CD115", "CD135", "CD150", "CXCR4", "ESAM", "CD4")
adtmat <- data[adtfeat, ]
allfeat <- rownames(data)
rnafeat <- setdiff(allfeat, adtfeat)
rnamat <- data[rnafeat, ]
rm(data, adtfeat, allfeat)
```

rm(adtmat, rnamat, rnafeat)
## Create Seurat Object
```{r create-seurat-object}
seurat_object <- CreateSeuratObject(counts = rnamat)
seurat_object[["ADT"]] <- CreateAssay5Object(counts = adtmat)

Assays(seurat_object)
seurat_object <- NormalizeData(object = seurat_object, assay = "RNA")
seurat_object <- NormalizeData(object = seurat_object, assay = "ADT", normalization.method = "RC")
rm(adtmat, rnamat, rnafeat)
```

Verify the features under each assay
```{r rownames, results='hide'}
rownames(seurat_object[["ADT"]])
rownames(seurat_object[["RNA"]])
```

Test ADT data by choosing any two ADT features
```{r featurescatter}
mpp3v2 <- SetIdent(mpp3v2, value = "ESAM")
Idents(cd135neg) <- "gates2"
FeatureScatter(cd135neg, feature1 = "adt_CD150", feature2 = "adt_CD48", log = TRUE) + labs(title="CD48 CD150")

DimPlot(cd135neg, group.by = "gates")
```


Check that the default assay is what you want it to be. Subsequent lines of code will use the features in the default assay unless you specify the assay within the arguments.
```{r default-assay}
# List the current default assay
DefaultAssay(seurat_object)
#Change default assay
DefaultAssay(seurat_object) <- "RNA" #change it to either "RNA" or "ADT"
DefaultAssay(seurat_object)
```

The Key for each assay allows you to put that before your feature of interest to circumvent switching the default assay (this is especially helpful if you want to use something from each assay in the same line of code).
```{r assay-key}
Key(seurat_object[["RNA"]])
Key(seurat_object[["ADT"]])
```


### Process the RNA data
```{r cluster-seurat-object, results='hide'}
seurat_object <- FindVariableFeatures(object = seurat_object)
seurat_object <- ScaleData(object = seurat_object)
seurat_object <- RunPCA(object =seurat_object)
seurat_object <- FindNeighbors(object =seurat_object, dims = 1:30)
seurat_object <- FindClusters(object =seurat_object)
seurat_object <- RunUMAP(object =seurat_object, dims = 1:30)
```

Display the UMAP
```{r dimplot}
DimPlot(mpp3v2, group.by = "ESAM")

FeaturePlot(seurat_object, features = "rna_ENSMUSG00000001946", min.cutoff = 0, max.cutoff = 1.5) +
  theme(legend.position = "none")
```
rna_ENSMUSG00000001946 = ESAM
# Analysis Tools

## Visualization
### Visualize both RNA and protein side by side
```{r featureplot}
DefaultAssay(seurat_object) <- "ADT"
p1 <- FeaturePlot(seurat_object, "ESAM", cols = c("lightgrey", "darkgreen")) + ggtitle("ESAM protein")
DefaultAssay(seurat_object) <- "RNA"
p2 <- FeaturePlot(seurat_object, "ENSMUSG00000001946") + ggtitle("ESAM RNA")
# place plots side-by-side
p1 | p2
```

An alternative way to write that code using the Key
```{r featureplot-v2}
p1 <- FeaturePlot(seurat_object, "adt_CD41", cols = c("lightgrey", "darkgreen")) + ggtitle("CD41 protein")
p2 <- FeaturePlot(seurat_object, "rna_ENSMUSG00000034664") + ggtitle("CD41 RNA")
p1|p2
```


### Create ADT scatter plots while showing cluster

Can "gate" cells using FeatureLocator
```{r featurescatter, eval=FALSE}
adtscatter <- FeatureScatter(seurat_object, feature1 = "adt_CD150", feature2 = "adt_CD135") +
                scale_x_log10() + 
                scale_y_log10()
adtscatter
seurat_object <- CellSelector(adtscatter, object = seurat_object, ident = "test")

```
limits = c(3, 500)
limits = c(20, NA)


Contour plot arguments: 
geom_point() +
geom_density_2d() +

geom_density_2d_filled()

Manual gating
``` {r mpp4 gating}
seurat_object$gates <- NULL

gating <- FetchData(seurat_object, vars = c("adt_CD150", "adt_CD135"))
gating_value <- gating[gating$adt_CD150 > 0 & gating$adt_CD135 > 0,]
gating_value <- gating_value[gating_value$adt_CD150 < 200,]
threshold <- quantile(gating_value$adt_CD135, 0.65)

seurat_object$gates <- ifelse(gating$adt_CD135 > threshold & gating$adt_CD150 < 200 & gating$adt_CD150 > 0, "MPP4", ifelse(gating$adt_CD135 < threshold & gating$adt_CD150 > 0 & gating$adt_CD135 > 0, "CD135-", NA))

```

```{r cd135- featurescatter}
Idents(cd135neg) <- "gates2"
adtscatter <- FeatureScatter(cd135neg, feature1 = "adt_CD150", feature2 = "adt_CD48") +
                scale_x_log10(limits = c(5, 10000)) + 
                scale_y_log10(limits = c(100, 10000))
                
adtscatter
cd135neg <- CellSelector(adtscatter, object = cd135neg, ident = "test")
```
limits = c(3, 800)
limits = c(1000, 10000)
 + 
                geom_density_2d(bins = 15)
                
``` {r}
max(FetchData(cd135neg, vars = "adt_CD150")[WhichCells(cd135neg, idents = "test"), "adt_CD150"])
```

Manual gating
``` {r gating}
cd135neg$gates2 <- NULL

gating <- FetchData(cd135neg, vars = c("adt_CD150", "adt_CD48"))

cd135neg$gates2 <- ifelse(gating$adt_CD48 > 4500 & gating$adt_CD150 < 200 & gating$adt_CD150 > 0, "MPP3", ifelse(gating$adt_CD48 < 4500 & gating$adt_CD150 > 0 & gating$adt_CD48 > 0 & gating$adt_CD150 < 200, "MPP1", ifelse(gating$adt_CD48 > 4500 & gating$adt_CD150 > 200, "MPP2", ifelse(gating$adt_CD48 < 4500 & gating$adt_CD48 > 0 & gating$adt_CD150 > 200, "HSC", cd135neg$gates))))


```

  ifelse(gating$adt_CD48 > 5600 & gating$adt_CD150 < 700 & gating$adt_CD150 > 0, "MPP3", ifelse(gating$adt_CD48 < 5600 & gating$adt_CD150 > 0 & gating$adt_CD48 > 0 & gating$adt_CD150 < 700, "MPP1", ifelse(gating$adt_CD48 > 2500 & gating$adt_CD150 > 700, "MPP2", ifelse(gating$adt_CD48 < 2500 & gating$adt_CD48 > 0 & gating$adt_CD150 > 700, "HSC", cd135neg$gates))))

```{r subsetting}
mpp4 <- subset(seurat_object, idents = "MPP4")
cd135neg <- subset(seurat_object, idents = "CD135-")
```


Verify
``` {r}
table(cd135neg$gates2)
```

Save gating
```{r save gating, eval=FALSE}
Idents(seurat_object) <- "seurat_clusters"
active_idents <- Idents(seurat_object)
gates <- c("MPP4", "CD135-")
seurat_object$gates <- ifelse(active_idents %in% gates, as.character(active_idents), NA)
```


To further gate, subset this gated population into a separate seurat object, gate the population of interest, then create a vector to add a new metadata column for these cells in the original object
```{r subset, eval=FALSE}
Idents(seurat_object) <- "gating"
mpp3v2 <- subset(seurat_object, idents = "CD135- MPP3")
```

```{r more gating, eval=FALSE}
subsetmpp3 <- FeatureScatter(mpp3v2, feature1 = "adt_CD150", feature2 = "adt_ESAM", log = TRUE)
mpp3v2 <- CellSelector(subsetmpp3, object = mpp3v2, ident = "ESAM+")
```
Save gating
```{r save gating, eval=FALSE}
active_idents <- Idents(mpp3v2)
gates <- c("ESAM+", "ESAM-")
mpp3v2$gate <- ifelse(active_idents %in% gates, as.character(active_idents), NA)
```

```{r revert gating, eval=FALSE}
modify_identifiers <- function(identifier) {
  if (identifier %in% c("CD135- MPP3 ESAM+", "CD135- MPP3 ESAM-", "CD135- MPP3 NA")) {
    return("CD135- MPP3")
  } else {
    return(identifier)
  }
}

seurat_object$gating <- sapply(seurat_object$gating, modify_identifiers)
```

ESAM+ vs - gating
```{r}
mpp3v2$gates <- NULL

esam <- FetchData(mpp3v2, vars = "adt_ESAM")
esam_value <- esam[esam > 0]
thirty <- quantile(esam_value, 0.3)
seventy <- quantile(esam_value, 0.7)
mpp3v2$ESAM <- NA
mpp3v2$ESAM[esam > seventy] <- "ESAM+"
mpp3v2$ESAM[esam < thirty & esam > 0] <- "ESAM-"

```


Attempt at a bar chart
```{r Bar Chart}
esam <- FetchData(seurat_object, vars = c("seurat_clusters", "adt_ESAM"))
esam <- data.frame(Cell = rownames(esam), Cluster = esam$seurat_clusters, ESAM = esam$adt_ESAM)
esam <- esam[order(esam$Cluster),]
esam$Cell <- factor(esam$Cell, levels = esam$Cell)

esambar <- ggplot(esam, aes(x = Cell, y = ESAM, fill = as.factor(Cluster))) + 
    geom_bar(stat = "identity", width = 1) +
  theme(
    axis.text = element_blank(), 
    axis.title = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )
 

cd9 <- FetchData(seurat_object, vars = c("seurat_clusters", "adt_CD9"))
cd9 <- data.frame(Cell = rownames(cd9), Cluster = cd9$seurat_clusters, CD9 = cd9$adt_CD9)
cd9 <- cd9[order(cd9$Cluster),]
cd9$Cell <- factor(cd9$Cell, levels = cd9$Cell)

cd9bar <- ggplot(cd9, aes(x = Cell, y = CD9, fill = as.factor(Cluster))) + 
    geom_bar(stat = "identity") + 
  theme(
    axis.text = element_blank(), 
    axis.title = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

barchart <- esambar / cd9bar

esambar

barchart  

```


```{r, eval=FALSE}
seurat_object$gating[Cells(mpp3v2)] <- paste("CD135- MPP3", mpp3v2$gate)
DimPlot(seurat_object, group.by = "gating")
```


### Similarly, to identify surface markers, you can make plots of ADT values the same way you can for RNA values.
```{r vlnplot}
VlnPlot(seurat_object, features = "rna_ENSMUSG00000045095", group.by = "seurat_clusters")
```

### View relationship between protein and RNA
```{r protein-v-rna}
FeatureScatter(seurat_object, feature1 = "adt_ESAM", feature2 = "rna_ENSMUSG00000001946")
```

### You can also FindMarkers for each assay
This block is currently written to identify markers in identifier 6.
```{r findmarkers, eval=FALSE}
adt_markers <- FindMarkers(seurat_object, ident.1 = "CD135- HSC ESAM+", ident.2 = "MPP4", assay = "ADT")
rna_markers <- FindMarkers(seurat_object, ident.1 = "CD135- HSC ESAM+", ident.2 = "MPP4", assay = "RNA")

write.csv(adt_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/adthscesam.csv")
write.csv(rna_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnahscesam.csv")

adt_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP1 ESAM+", ident.2 = "MPP4", assay = "ADT")
rna_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP1 ESAM+", ident.2 = "MPP4", assay = "RNA")

write.csv(adt_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/adtmpp1esam.csv")
write.csv(rna_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp1esam.csv")

adt_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP2 ESAM+", ident.2 = "MPP4", assay = "ADT")
rna_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP2 ESAM+", ident.2 = "MPP4", assay = "RNA")

write.csv(adt_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/adtmpp2esam.csv")
write.csv(rna_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp2esam.csv")

adt_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP3 ESAM+", ident.2 = "MPP4", assay = "ADT")
rna_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP3 ESAM+", ident.2 = "MPP4", assay = "RNA")

write.csv(adt_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/adtmpp3esam.csv")
write.csv(rna_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3esam.csv")

adt_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP3 ESAM+", ident.2 = "CD135- MPP3 ESAM-", assay = "ADT")
rna_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP3 ESAM+", ident.2 = "CD135- MPP3 ESAM-", assay = "RNA")

write.csv(adt_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/adtmpp3v3.csv")
write.csv(rna_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3v3.csv")
```

```{r}
adt_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP3 ESAM+", ident.2 = "CD135- MPP3 ESAM-", assay = "ADT")
rna_markers <- FindMarkers(seurat_object, ident.1 = "CD135- MPP3 ESAM+", ident.2 = "CD135- MPP3 ESAM-", assay = "RNA")

write.csv(adt_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/adtmpp3.csv")
write.csv(rna_markers, "C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3.csv")

EnsIDReplace("C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3.csv", "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3_genename.csv")
```

```{r ESAM DEG GSEA}
adt_markers <- FindMarkers(mpp3v2, ident.1 = "ESAM+", ident.2 = "ESAM-", assay = "ADT")
rna_markers <- FindMarkers(mpp3v2, ident.1 = "ESAM+", ident.2 = "ESAM-", assay = "RNA")

write.csv(adt_markers, "adtmpp3.csv")
write.csv(rna_markers, "rnampp3.csv")

EnsIDReplace2(rna_markers, mpp3esam)
write.csv(mpp3esam_newgene, "rnampp3_genename.csv")

run_gsea(mpp3esam_newgene, "reactome_go_gsea.gmt", "mpp3esam_gsea.csv")
```

Replace RNA Ensembl IDs with gene IDs (this part is still buggy, leaves some blank cells, and uses 40 gb of RAM. So skip it). It works just fine if you do it for the files generated by FindMarkers

```{r Ensembl ID Replacement, eval=FALSE}
EnsIDReplace <- function(input, output){
rna <- as.sparse(read.csv(file = input, header = TRUE, row.names = 1))
rnanames <- rownames(rna)
ensembl <- useEnsembl(biomart = "genes")
listDatasets(ensembl)
mart <- useDataset("mmusculus_gene_ensembl", useMart("ensembl"))
gene_ids <- getBM(
  filters = "ensembl_gene_id",
  attributes = c("ensembl_gene_id", "mgi_symbol"),
  values = rnanames,
  mart = mart
)

id_list <- split(gene_ids[["mgi_symbol"]], gene_ids[["ensembl_gene_id"]])
rnaname <- character(nrow(rna))
for (i in seq_len(nrow(rna))) {
  ensembl_id <- rownames(rna)[i]
  if (ensembl_id %in% names(id_list)) {
    rnaname[i] <- id_list[[ensembl_id]]
  } else{
    rnaname[i] <- ensembl_id
  }
}
rnaname[rnaname == ""] <- NA

rownames(rna) <- rnaname
rm(ensembl, gene_ids, id_list, mart, ensembl_id, i, rnaname)
write.csv(rna, output)
}

EnsIDReplace("C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnahscesam.csv", "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnahscesam_genename.csv")

EnsIDReplace("C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp1esam.csv", "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp1esam_genename.csv")

EnsIDReplace("C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp2esam.csv", "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp2esam_genename.csv")

EnsIDReplace("C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3esam.csv", "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3esam_genename.csv")

EnsIDReplace("C:/Users/rthalla//OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3v3.csv", "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/RNASeq/Cite-seq/GSE145491/rnampp3v3_genename.csv")
```
For further analysis it can be easier to define the genes you want to use ahead of time
```{r}
MkPspecific <- c("Maml2", "Mtss1l","Slc39a13", "Rell2", "Nrgn", "S100a1", "Lrch4", "Rnasek", "Gucy1b3")
Meglin <- c("Selp", "Itga2b", "Itgb3", "Cd9")
Megcomb <- c("Maml2", "Mtss1l","Slc39a13", "Rell2", "Nrgn", "S100a1", "Lrch4", "Rnasek", "Gucy1b3", "Selp", "Itga2b", "Itgb3", "Cd9")
meg <- c("Apoe", "Vwf")
coag <- c("Timp3", "Pf4", "Vwf", "Gda", "Gp1ba", "Itgb3", "F2rl2", "Pdgfb", "P2ry1", "Pros1", "Cd9", "Gp9", "Mmp14", "Clu", "Fn1", "Maff", "S100a1", "Bmp1", "Thbd")
```

```{r}
ensembl <- useEnsembl(biomart = "genes")
listDatasets(ensembl)
mart <- useDataset("mmusculus_gene_ensembl", useMart("ensembl"))
gene_ids <- getBM(
  filters = "mgi_symbol",
  attributes = c("mgi_symbol", "ensembl_gene_id"),
  values = coag,
  mart = mart
)

id_list <- split(gene_ids[["ensembl_gene_id"]], gene_ids[["mgi_symbol"]])
CoagEns <- gene_ids[["ensembl_gene_id"]]

rm(ensembl, gene_ids, id_list, mart)
```


Generate heatmap
```{r seurat heatmap}
DefaultAssay(seurat_object) <- "RNA"
DoHeatmap(seurat_object, features = CoagEns, group.by = "gating")
```
 "CD135- HSC ESAM-", "CD135- MPP1 ESAM-","CD135- MPP2 ESAM-", "CD135- MPP3 ESAM-",

```{r pheatmap}
expression_data <- FetchData(seurat_object, vars = CoagEns)
expression_data$gating <- seurat_object@meta.data$gating

# Filter data to include only the desired groups
desired_groups <- c("CD135- HSC ESAM+", "CD135- MPP1 ESAM+", "CD135- MPP2 ESAM+",  "CD135- MPP3 ESAM+", "MPP4")
filtered_data <- expression_data[expression_data$gating %in% desired_groups, ]

# Prepare heatmap data
heatmap_data <- as.matrix(filtered_data[, CoagEns])
rownames(heatmap_data) <- rownames(filtered_data)
heatmap_data <- heatmap_data[order(filtered_data$gating), ]

rnanames <- colnames(heatmap_data)
ensembl <- useEnsembl(biomart = "genes")
listDatasets(ensembl)
mart <- useDataset("mmusculus_gene_ensembl", useMart("ensembl"))
gene_ids <- getBM(
  filters = "ensembl_gene_id",
  attributes = c("ensembl_gene_id", "mgi_symbol"),
  values = rnanames,
  mart = mart
)

id_list <- split(gene_ids[["mgi_symbol"]], gene_ids[["ensembl_gene_id"]])
rnaname <- character(ncol(heatmap_data))
for (i in seq_len(ncol(heatmap_data))) {
  ensembl_id <- colnames(heatmap_data)[i]
  if (ensembl_id %in% names(id_list)) {
    rnaname[i] <- id_list[[ensembl_id]]
  } else{
    rnaname[i] <- ensembl_id
  }
}
rnaname[rnaname == ""] <- NA

colnames(heatmap_data) <- rnaname
rm(ensembl, gene_ids, id_list, mart, ensembl_id, i, rnaname)

# Define color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(100)

# Create a heatmap
pheatmap(heatmap_data, 
         cluster_rows = FALSE, 
         cluster_cols = TRUE, 
         color = color_palette,
         show_rownames = FALSE,
         annotation_row = filtered_data["gating"],
         main = "Heatmap of Meg/MkP Genes")
```

# Create a heatmap
pheatmap(seurat_object, 
         scale = "row", 
         clustering_distance_rows = "euclidean", 
         cluster_cols = FALSE, 
         clustering_method = "complete", 
         color = color_palette,
         show_rownames = TRUE, 
         show_colnames = TRUE,
         main = "Heatmap of AITL Genes")