---
title: "MPP3 ESAM RNA Sequencing Analysis"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

# Initial Setup:

Load required packages (this is assuming you already have these packages installed). Always run this first.

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(DESeq2)
library(rnaseqGene)
library(pheatmap)
library(ggalt)
library(limma)
library(sva)
```

# DESeq2 analysis

## Load counts matrix into a DESeqDataSet object

```{r Load MPP3 Data, warning=FALSE, message=FALSE}
mpp3 <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/MPP3 RNA Seq/newgenenames.csv", row.names = 1) #file path to Counts csv document
```

```{r Load MPC Data, warning=FALSE, message=FALSE}
cts2 <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/MPP3 RNA Seq/MPC_Counts_Corrected.csv") #file path to Counts csv document
cts2 <- cts2[, -1] #only necessary if bulk seq data has ensembl id before gene names
cts <- cts2[, -1]
table(duplicated(cts$Gene.name)) #check for duplicate gene names
rownames(cts) <- make.names(cts2$Gene.name, unique = TRUE) #forces gene names to become unique
rm(cts2)
mpc <- cts
rm(cts)
```

```{r Combine Batches, warning=FALSE, message=FALSE}
common_genes <- intersect(rownames(mpp3), rownames(mpc))
counts1 <- mpp3[common_genes, ]
counts2 <- mpc[common_genes, ]
cts <- cbind(counts1, counts2)
rm(counts1, counts2, common_genes)
```

## Differential Expression analysis

```{r DESeq}
cd <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/MPP3 RNA Seq/coldata.csv") #make coldata file with sample names in one column and then condition in second column
dds <- DESeqDataSetFromMatrix(countData = cts, colData = cd, design = ~condition)
rlog <- rlog(dds, blind=FALSE)
dds <- DESeq(dds)
rescmp <- results(dds) 
```

### Save results as csv

```{r Save Results, eval=FALSE}
write.csv(rescmp, "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/MPP3 RNA Seq/completeresults.csv")
```

# Generate Heatmap

This heatmap is generated using a list of AITL genes curated using other datasets. 

```{r Heatmap of Specific Genes}
# Subset normalized counts to include only the genes of interest
subset_rlog <- assay(rlog)[rownames(rlog) %in% overlap, ]
dko_samples <- c("DKO_1", "DKO_2", "DKO_3")
rest <- c("WT_1", "WT_2", "WT_3", "Tet2KO_1", "Tet2KO_2", "Tet2KO_3")
row_order <- order(rowMeans(subset_rlog[, dko_samples])-rowMeans(subset_rlog[, rest]), decreasing = TRUE)
subset_rlog <- subset_rlog[row_order, ]
subset_rlog <- as.matrix(subset_rlog)

# Define the desired column order
desired_order <- c("WT_1", "WT_2", "WT_3", "Tet2KO_1", "Tet2KO_2", "Tet2KO_3", "DKO_1", "DKO_2", "DKO_3")

# Reorder the columns
subset_rlog <- subset_rlog[, desired_order]

# Extract column data (sample information) and reorder it
col_data <- colData(dds)[desired_order, ]

# Define color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(100)

# Create a heatmap
heatmap <- pheatmap(subset_rlog, 
         scale = "row", 
         clustering_distance_rows = "binary", 
         cluster_cols = FALSE, 
         clustering_method = "complete", 
         color = color_palette,
         show_rownames = TRUE, 
         show_colnames = TRUE,
         main = "Heatmap of AITL Genes",
         treeheight_row = 0)

```

# Generate PCA Plot

```{r PCA Plot}
pca <- plotPCA(rlog)
pca + geom_point(size=5) + 
  theme(
    panel.background = element_rect(fill = "white", color = "black"), # White background
    axis.text = element_blank(),
    axis.ticks = element_blank(),
  ) + 
  scale_color_manual(labels=c("MPP3 ESAM-", "MPP3 ESAM+", "GMP", "PG", "CMP", "PM", "MEP"), values = c("red","blue", "darkgreen", "purple", "orange", "black", "lightblue"))

```
  lims(x=c(-15, 25), y=c(-15, 25))
# Display GSEA Results

## Filter GSEA reports for an FDR q-value < 0.25 and Normalized Enrichment Score > 1.3

The negative reports were not added for the DKO vs WT comparison as the report was empty, and that created some errors. 

```{r GSEA Results}
dt1 <- read_tsv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/newaitl2.Gsea.1728424109372/gsea_report_for_DKO_1728424109372.tsv")
dt2 <- read_tsv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/newaitl2.Gsea.1728424109372/gsea_report_for_Tet2KO_1728424109372.tsv")
dt <- bind_rows(dt1, dt2)
rm(dt1, dt2)
dt <- dt[which(dt["FDR q-val"]<0.25 & abs(dt["NES"])>1.2),]
dt <- rename(dt, FDR.q.val="FDR q-val")

write.csv(dt, "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/filtereddkotet2.csv")

dw <- read_tsv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/newaitl6.Gsea.1728428240183/gsea_report_for_DKO_1728428240183.tsv")
dw <- dw[which(dw["FDR q-val"]<0.25 & abs(dw["NES"])>1.2),]
dw <- rename(dw, FDR.q.val="FDR q-val")

write.csv(dw, "C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/filtereddkowt.csv")
```


## Generate Plots

The filtered reports from above were then separated into three categories to then be plotted together. 

#### DKO vs Tet2KO comparison results
```{r Generate Plots - DKO vs Tet2KO, fig.width=10, fig.height=9}
tf <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/finaltfolldtv2.csv")
tfplot <- ggplot(tf, aes(x=-log10(FDR.q.val+0.00001), y=reorder(NAME, -log10(FDR.q.val)), fill = NES))
tfplot + geom_point(size=5, shape=21) + 
 theme(
    panel.background = element_rect(fill = "white"), 
    panel.border = element_rect(color = "black", fill = NA), 
    plot.title = element_text(hjust = 0.5), 
    axis.text=element_text(color="black", face="bold")
    ) +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0, limits=c(-2.1,2.1)) + 
  labs(x="-log(FDR q value)", y="Genesets", title="T cell & Lymphoma Genesets") +
  xlim(0,5.1)

cc <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/finalcellcycledtv2.csv")
ccplot <- ggplot(cc, aes(x=-log10(FDR.q.val+0.00001), y=reorder(NAME, -log10(FDR.q.val)), fill = NES))
ccplot + geom_point(size=5, shape=21) +  
 theme(
    panel.background = element_rect(fill = "white"), 
    panel.border = element_rect(color = "black", fill = NA), 
    plot.title = element_text(hjust = 0.5), 
    axis.text=element_text(color="black", face="bold")
    ) +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0, limits=c(-2.2,2.2)) + 
  labs(x="-log(FDR q-value)", y="Genesets", title="Cell Cycle & Proliferation Genesets") +
  xlim(0,5.1)

my <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/finalmyeloiddtv2.csv")
myplot <- ggplot(my, aes(x=-log10(FDR.q.val+0.00001), y=reorder(NAME, -log10(FDR.q.val)), fill = NES))
myplot + geom_point(size=5, shape=21) +  
 theme(
    panel.background = element_rect(fill = "white"), 
    panel.border = element_rect(color = "black", fill = NA), 
    plot.title = element_text(hjust = 0.5), 
    axis.text=element_text(color="black", face="bold")
    ) +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0, limits=c(-2.0,2.0)) + 
  labs(x="-log(FDR q-value)", y="Genesets", title="Myeloid Genesets") +
  xlim(0,5.1)
```

#### DKO vs WT comparison results
```{r Generate Plots - DKO vs WT, fig.width=10, fig.height=9}
tf <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/finaltfolldwv2.csv")
tfplot <- ggplot(tf, aes(x=-log10(FDR.q.val+0.00001), y=reorder(NAME, -log10(FDR.q.val)), fill = NES))
tfplot + geom_point(size=5, shape=21) + 
 theme(
    panel.background = element_rect(fill = "white"), 
    panel.border = element_rect(color = "black", fill = NA), 
    plot.title = element_text(hjust = 0.5), 
    axis.text=element_text(color="black", face="bold")
    ) +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0, limits=c(-1.5,1.5)) + 
  labs(x="-log(FDR q value)", y="Genesets", title="T cell & Lymphoma Genesets") +
  xlim(0,5.1)

cc <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/finalcellcycledwv2.csv")
ccplot <- ggplot(cc, aes(x=-log10(FDR.q.val+0.00001), y=reorder(NAME, -log10(FDR.q.val)), fill = NES))
ccplot + geom_point(size=5, shape=21) +  
 theme(
    panel.background = element_rect(fill = "white"), 
    panel.border = element_rect(color = "black", fill = NA), 
    plot.title = element_text(hjust = 0.5), 
    axis.text=element_text(color="black", face="bold")
    ) +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0, limits=c(-1.7,1.7)) + 
  labs(x="-log(FDR q-value)", y="Genesets", title="Cell Cycle & Proliferation Genesets") +
  xlim(0,5.1)

my <- read.csv("C:/Users/rthalla/OneDrive - Loyola University Chicago/Zhang Lab/Kanak Tet2 Ripk3/output/202410083/finalmyeloiddwv2.csv")
myplot <- ggplot(my, aes(x=-log10(FDR.q.val+0.00001), y=reorder(NAME, -log10(FDR.q.val)), fill = NES))
myplot + geom_point(size=5, shape=21) +  
 theme(
    panel.background = element_rect(fill = "white"), 
    panel.border = element_rect(color = "black", fill = NA), 
    plot.title = element_text(hjust = 0.5), 
    axis.text=element_text(color="black", face="bold")
    ) +
  scale_fill_gradient2(low="blue", mid="white", high="red", midpoint=0, limits=c(-1.6,1.6)) + 
  labs(x="-log(FDR q-value)", y="Genesets", title="Myeloid Genesets") +
  xlim(0,5.1)
```
